Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Pass at trying to jazz up intro

\section*{INTRODUCTION}

%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
%%% and nonstandard abbreviations should be defined when first used in the text.

The genomic revolution has transformed how we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation,
assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
the tree-like representation of UShER is no longer tenable.

@@

% TODO merge this in
Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as ``robust'' recombination events,
Figure~\ref{fig:recombinant_qc}; see STAR Methods)
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass the robustness filtering criteria (Figure~\ref{fig:recombinant_qc}).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with the XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

Furthermore, we compared the sensitivity of sc2ts and rebar.
We ran rebar on the ``robust'' recombination events detected by sc2ts (STAR Methods),
and found that 167 (43\%) of them were also detected as recombinants by rebar.
We hypothesize that this disagreement arose from
differences in sensitivity between these methods.
Consistent with this, we found that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).

@@

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
\label{fig:averted_mutations}
\end{figure}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous studies \cite{Turakhia2022,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination intervals.
(A) Distribution of the breakpoint intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
\label{fig:recombinant_intervals}
\end{figure}

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 356 ``robust'' events.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

\subsubsection*{Quality control of recombination events}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labeled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

>>>>

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with 
% Not sure about this assertion! Yan?
samples separated by millions of years of evolution.
[This is rough outline text]
Complex models are required to fill in the gaps in our knowledge,
incorporating prior information about sequence evoltion etc etc.
In this context, recombination can be treated as a nuisance factor 
because we cannot hope to observe it over these timescales [or something
similar].
Phylogenetics has continued along this vein, 
although with various challenges on how to accomodate 
more closely related samples and to handle potential recombination met.

The data collected during the COVID-19 pandemic as part of global
surveillance fundamentally changed these assumptions. 
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. This superabundance of data 
overwhelmed classical phylogenetic methods, both in terms 
of the number of sequences that could be processed, but also
more fundamentally in terms of the underlying assumption of 
retrospective data sparsity. [TODO rephrase]
The UShER method \cite{Turakhia2021} conclusively demonstrated 
the value of simpler models that build a tree iteratively
in real-time, quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.

Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Despite this, phylogenies used as the basis for innumerable studies 
continue to depict the genealogy of SARS-CoV-2 as a single
tree~\cite{McBroome2021,Hadfield2018}.

% The genomic revolution has transformed how we study the evolution and spread of pathogens,
% allowing real-time tracking of pathogens over time and space.
% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
% quickly overwhelming typical phylogenetic approaches.
% Although real-time phylogenetic methods now exist to cope with this flood of genomic data

% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
% we lack methods to build phylogenies that can integrate recombination
% in the face of ongoing sampling of infected cases.

% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
% UShER has generated phylogenies updated daily \cite{McBroome2021} and
% has proven to be a powerful resource that successfully enabled the world
% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
% These phylogenies are used to identify and monitor
% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
% which might have acquired features to become Variants of Concern (VOC),
% for example, increased transmissibility, disease severity, and immune escape.
% Beyond supporting real-time genomic epidemiology,
% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
% UShER uses a tree to represent evolutionary change via mutation,
% assuming no recombination \cite{Turakhia2021}.

@@

@@

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 354 pass quality control (QC) filters
(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass QC (STAR Methods).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B), and which we tentatively group using the unused label Xx.
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
believe the original recombinant designation to be based on erroneous data
(Supplementary document S1.XXX).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.x and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

% https://github.com/jeromekelleher/sc2ts-paper/issues/717
167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
We hypothesize that this disagreement arose from
differences in sensitivity between these methods.
Consistent with this, we found that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).

@@

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, Xx 
(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous studies
\cite{Bobay2020recombination,Turakhia2022,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination intervals.
(A) Distribution of the breakpoint intervals of 354 recombination events 
along the genome (annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
\label{fig:recombinant_intervals}
\end{figure}

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 354 events passing QC filters.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

\subsubsection*{Quality control of recombination events}
Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect can be identified
because they affect multiple adjacent or near-adjacent bases simultaneously.
